---
title: "Getting started with gwasplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with gwasplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 4.5,
  dpi = 96,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(gwasplot)
```

**gwasplot** turns a tidy table of GWAS summary statistics into
publication-ready and interactive figures. This vignette walks through the
data contract, the three plot types, top-marker highlighting, interactivity,
and the top-hits table.

## The data contract

Every function expects the four canonical columns `SNP`, `CHR`, `POS` and `P`
(plus optional `gene` and `trait`). You rarely have to prepare them by hand ---
`validate_gwas()` auto-detects common column-name aliases and coerces types:

```{r}
raw <- data.frame(
  marker = c("rs1", "rs2", "rs3"),
  chrom  = c(1, 1, 2),
  bp     = c(100, 200, 150),
  pvalue = c(1e-8, 0.2, 3e-3)
)
validate_gwas(raw)
```

All the plotting functions call `validate_gwas()` internally, so you can pass
raw data directly. Two simulated datasets ship with the package:

```{r}
data(gwas_example)  # single trait, 22 chromosomes
data(gwas_multi)    # three traits: height, bmi, glucose
gwas_example
```

## Manhattan plot

```{r manhattan}
gwas_manhattan(gwas_example)
```

Chromosomes are laid end-to-end with alternating colour bands; the solid line
is the genome-wide threshold (`5e-8`) and the dashed line is the suggestive
threshold (`1e-5`).

## Highlighting top markers

`highlight_top()` bundles a selection rule that is shared by every plot *and*
the table, so a figure and its table always agree. Combine any of `top_n`,
`threshold`, `snps`, `genes`, and `by`:

```{r manhattan-highlight}
gwas_manhattan(gwas_example, highlight = highlight_top(top_n = 6))
```

`by` groups before taking `top_n` --- for example, the lead SNP on every
chromosome:

```{r manhattan-bychr}
gwas_manhattan(gwas_example, highlight = highlight_top(top_n = 1, by = "CHR"),
               label = FALSE)
```

## Zooming into one chromosome

```{r chromosome, fig.width = 7}
gwas_chromosome(gwas_example, chr = 1, highlight = highlight_top(top_n = 3))
```

## Circular plots with multiple rings

`gwas_circular()` arranges the chromosomes once around a circle and draws each
trait as its own concentric ring (outermost = first trait). This is the
CMplot-style view.

```{r circular, fig.width = 7, fig.height = 6.5}
gwas_circular(gwas_multi, highlight = highlight_top(top_n = 1, by = "trait"))
```

Useful options:

* `shared_scale = FALSE` scales each ring to its own maximum (peaks fill the
  ring) instead of a common scale.
* `ideogram = FALSE` drops the central chromosome band.
* `colors` sets one colour per ring.

```{r circular-opts, fig.width = 7, fig.height = 6.5}
gwas_circular(gwas_multi, shared_scale = FALSE, ideogram = FALSE,
              colors = c("#0b6e4f", "#e28413", "#8367c7"))
```

A single-trait frame collapses to one ring automatically:

```{r circular-single, fig.width = 6.5, fig.height = 6.5}
gwas_circular(gwas_example, highlight = highlight_top(top_n = 1, by = "CHR"))
```

> **Tip.** When you highlight many markers (e.g. every genome-wide hit via a
> `threshold`), set `label = FALSE` to avoid a wall of overlapping labels.

## Interactive plots

Pass `interactive = TRUE` to any of the three plots to turn it into a
[ggiraph](https://davidgohel.github.io/ggiraph/) htmlwidget. Every point then
carries a hover tooltip with its **SNP, gene, chromosome, position and
p-value**; you can also **scroll to zoom** and **drag to pan**, and hovering
highlights the point. This works in HTML output such as this vignette, an R
Markdown / Quarto report, or a Shiny app.

### Interactive Manhattan

Hover any peak to read off the exact marker behind it --- no need to squint at
labels or cross-reference a table. Because an interactive plot turns every point
into an SVG node, we first **thin** the data: keep all the interesting hits plus
a small random sample of the null background. This keeps the widget snappy while
leaving every peak intact.

```{r int-manhattan, eval = requireNamespace("ggiraph", quietly = TRUE), fig.width = 8}
set.seed(1)
keep <- gwas_example$P < 1e-3 | runif(nrow(gwas_example)) < 0.15
gwas_manhattan(gwas_example[keep, ], highlight = highlight_top(top_n = 6),
               interactive = TRUE)
```

### Interactive circular

The tooltips are per ring, so a hovered point also tells you **which trait** it
belongs to. Here we use a small simulated dataset so the widget stays light ---
interactive SVGs grow with the number of points, so prefer to thin very dense
data first (see the note below).

```{r int-circular, eval = requireNamespace("ggiraph", quietly = TRUE), fig.width = 7, fig.height = 6.5}
demo <- simulate_gwas(n_chr = 12, snps_per_chr = 120,
                      traits = c("trait A", "trait B"), seed = 11)
gwas_circular(demo, highlight = highlight_top(top_n = 1, by = "trait"),
              interactive = TRUE)
```

### Interactive single chromosome

Zooming into a region and hovering is a quick way to inspect a locus:

```{r int-chrom, eval = requireNamespace("ggiraph", quietly = TRUE), fig.width = 7}
gwas_chromosome(gwas_example, chr = 1, highlight = highlight_top(top_n = 3),
                interactive = TRUE)
```

## Very large studies (millions of SNPs)

A genome-wide study can carry tens of millions of SNPs, which overwhelms
ggplot2 (one grob per point) and browsers (one SVG node per point). Set
`big_data = TRUE` to switch on a pipeline built for this:

* the dense **background is rasterised** into a bitmap layer (via
  [scattermore](https://github.com/exaexa/scattermore)) --- *all* points are
  kept, but render time and file size stay bounded;
* the **significance lines and highlighted markers stay crisp vectors** on top;
* when combined with `interactive = TRUE`, only the **highlighted markers**
  become an interactive layer (they are the only points worth hovering), so the
  widget never has to hold millions of nodes.

Here is a real example. `huge` has about a quarter of a million SNPs; on your
own data you would simply pass a much larger frame. Because the background is a
bitmap, the figure renders almost instantly and stays a small PNG even though
every point is drawn --- the ten strongest markers sit on top as crisp,
labelled vector points.

```{r big-data, fig.width = 8, fig.height = 4.2}
huge <- simulate_gwas(n_chr = 22, snps_per_chr = 18000, n_peaks = 20, seed = 5)
nrow(huge)
gwas_manhattan(huge, big_data = TRUE, highlight = highlight_top(top_n = 10))
```

For an interactive version, add `interactive = TRUE`: the background stays a
raster while only the highlighted markers carry tooltips, so the widget never
has to hold millions of nodes.

```{r big-data-int, eval = FALSE}
gwas_manhattan(huge, big_data = TRUE, interactive = TRUE,
               highlight = highlight_top(top_n = 10))
```

`big_data = FALSE` (the default) leaves the ordinary vector plot completely
unchanged. If you would rather reduce the data yourself --- for a table, or to
pass a smaller frame around --- [thin_gwas()] keeps every hit and grid-samples
the null background down to a target size:

```{r thin}
small <- thin_gwas(huge, max_points = 5000)
nrow(small)
```

## Top-markers table

`gwas_top()` returns the selected markers as a plain tibble, and `gwas_table()`
formats them as a [gt](https://gt.rstudio.com) table --- positions with
thousands separators, p-values in scientific notation, and a colour-scaled
$-\log_{10}(P)$ column.

```{r top}
gwas_top(gwas_example, highlight = highlight_top(top_n = 5))
```

```{r table, eval = requireNamespace("gt", quietly = TRUE)}
gwas_table(gwas_example, highlight = highlight_top(top_n = 8),
           title = "Top GWAS hits")
```

Because the table and the plot share `highlight_top()`, you can highlight the
same markers in both:

```{r paired, eval = FALSE}
hl <- highlight_top(top_n = 8)
gwas_manhattan(gwas_example, highlight = hl)   # figure
gwas_table(gwas_example, highlight = hl)        # matching table
```
